Reducing RES Droughts through the integration of wind and solar PV

### 2025-04-02 Article/RES_droughts_Morin_ReEn_R3.tex
Update Reviewed Version of the Article

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{violin_plots_verification.pdf}
	\caption{Violin plots of CF distributions for wind a) and solar PV b) for the Observations (grey) and the three datasets: ATL (red), C3S~GRD (blue), and C3S~NAT (purple). The black dot shows the median values, while the black vertical lines represent the first and third quartiles. The red line indicates the threshold of 0.1 used in the study to identify RES droughts}
	\label{fig:violin_plots}
\end{figure}

@@

This study aimed to answer two key questions: Do generic datasets have sufficient skill to reliably quantify RES drought events? How does the integration of solar PV into a predominantly wind-based system alter the characteristics of RES droughts? To address these questions, three datasets were compared: one developed using a regionally validated model and two derived from the European C3S-Energy dataset. The C3S-Energy datasets differ in their assumptions, one assumes a homogeneous distribution of wind and solar PV across the region, while the other includes the actual locations of RES farms. In contrast, the regionally validated model accounts for farm locations and uses tailored wind and solar PV models selected to represent the actual generation.

Our results demonstrate that datasets without regional validation have notable limitations in quantifying RES drought events. While all three datasets capture broad trends, significant differences arise in extreme event assessment. In particular, datasets without regional validation tend to underestimate the return periods of RES droughts, which can lead to insufficient reserve capacity planning and potential risks to grid stability and security of supply. This highlights the critical need for regionally validated datasets when studying RES droughts in renewable energy planning. The inclusion of wind and solar PV farm locations has a limited impact on RES drought analysis compared to the choice of wind turbine power curves and solar PV models.

The effect of the integration of solar PV in a wind-dominated system on RES droughts has been explored. Our analysis has demonstrated that transitioning to a system with more equal amounts of wind and solar PV capacity reduces the frequency, duration, and seasonal variability of RES drought events. This improvement is attributed to the complementary nature of wind and solar PV generation, as solar PV generation typically peaks in summer while wind generation is more consistent in winter. However, this integration is unable to counter the critical winter RES droughts, which coincide with the strongest electricity demand in Northern European countries. Still, a more diversified renewable energy mix mitigates extreme RES drought conditions and enhances overall system resilience.

The results presented in this study have four main limitations. First, the presented study uses a fixed threshold to define RES drought events, but other methods such as ones using the percentile of the generation could yield different results. Second, the definition of RES droughts based on generation does not consider the important role of demand, which could be of interest to system operators. Third, recent solar PV capacity expansions have changed the generation profile, limiting solar PV data for model training to a single year, whereas a longer validation period would be preferable. Lastly, the source for weather data is ERA5, which is among the best reanalysis datasets for renewable energy applications, but still comes in a limited spatial resolution, an issue that can be addressed once higher resolution datasets become available.

>>>>

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{violin_plots_verification.pdf}
	\caption{Violin plots of CF distributions for a)  wind and b) solar PV for the Observations (grey) and the three datasets: ATL (red), C3S~GRD (blue), and C3S~NAT (purple). The black dot shows the median values, while the black vertical lines represent the first and third quartiles. The black dashed line indicates the threshold of 0.1 used in the study to identify RES droughts}
	\label{fig:violin_plots}
\end{figure}

@@

This study aimed to answer two key questions: Do generic datasets have sufficient skill to reliably quantify RES drought events? How does the integration of solar PV into a predominantly wind-based system alter the characteristics of RES droughts? To address these questions, three datasets were compared: two derived from the European C3S-Energy dataset, and one developed by the authors. The datasets derived from C3S-Energy differ in their assumptions, one assumes a homogeneous distribution of wind and solar PV capacity across the region, while the other includes the actual locations of RES farms. The dataset developed by the authors uses a regionally validated model which accounts for farm locations and uses tailored wind and solar PV models selected to represent the actual generation.

Our results demonstrate that datasets without regional validation misrepresent the frequency and duration of RES drought events due to their limited ability to reproduce the observations. The inclusion of wind and solar PV farm locations has limited impact on RES drought analysis compared to the choice of wind turbine power curves and solar PV models. Whereas all three datasets capture broad trends in the duration and seasonality of RES drought events, the actual number of events is consistently underestimated by the non-validated datasets. This effect becomes clearer for extreme events, as not using regionally validated datasets can yield an overestimation of the return periods of RES droughts. This can lead to insufficient reserve capacity planning and potential risks to grid stability and security of supply.

The effect of the integration of solar PV capacity in a wind-dominated system on RES droughts has been explored. Our analysis has demonstrated that transitioning to a system with more equal amounts of wind and solar PV capacity reduces the occurrence of RES drought events, mitigates extreme RES drought conditions and enhances overall system resilience. This improvement is attributed to the complementary nature of wind and solar PV generation, as solar PV generation typically peaks in summer while wind generation predominates during winter. However, this integration is unable to counter the critical winter RES droughts, which coincide with the strongest electricity demand in Northern European countries.

The results presented in this study have four main limitations. First, the presented study uses a fixed threshold to define RES drought events, but other methods such as ones based on percentiles could be explored. Second, the definition of RES droughts based on generation does not consider the important role of demand, which could be of interest to system operators. Third, recent solar PV capacity expansions have changed the generation profile, limiting solar PV data for model training to a single year, although a longer validation period would be preferable. Last, the source for weather data is ERA5 has limited spatial resolution, an issue that can be addressed once higher resolution datasets become available.
